MS Technical Paper: \\ Placement Algorithms for Heterogenous FPGAs

### 2025-03-31 latex/paper/paper.tex
more paperwork

\documentclass[twocolumn]{article}
\usepackage{hyperref}
\usepackage{enumitem}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{mathpazo}
% \usepackage{multicol}
\usepackage[a4paper,
            bindingoffset=0.2in,
            left=1in,
            right=1in,
            top=1in,
            bottom=1in,
            footskip=.25in]{geometry}

@@

    This brings us to modern day FPGA architectures.
    To meet the needs of increasingly complex designs, FPGA vendors introduced heterogeneous FPGAs. 
    In these devices, hard macros such as Block RAM (BRAM) and Digital Signal Processing (DSP) slices are integrated into the programmable logic fabric along with CLBs. 
    This design enables the direct instantiation of common subsystems like memories and multipliers, without having to implement them solely with CLBs. 
    Major vendors such as Xilinx and Altera now employ heterogeneous island-style architectures in their devices. 
    As designs become increasingly large and complex, FPGAs become increasingly heterogenous as they incorporate a wider variety of hard macros into the fabric.

@@

    \begin{figure}
        \centering
        \includegraphics[width=7.0cm]{figures/design_flow.png}
        \caption{A typical FPGA design and verification flow}
        \label{fig:design_flow}
    \end{figure}

    Figure \ref{fig:design_flow} shows a typical FPGA design and verification flow.
    The FPGA design flow takes a high-level digital design written in an HDL and generates a configuration file that programs the physical FPGA.

@@

    Next, the \textbf{synthesis} tool interprets the HDL code and generates a netlist that represents the design in terms of basic logic elements, such as lookup tables (LUTs) and flip flops. 
    During this stage, substages like instantiation, substitution, inference, optimization, and technology mapping are applied to map the user design to a particular FPGA architecture.

    After synthesis, the \textbf{placement} tool assigns each logic element in the netlist a specific location within the FPGA’s fabric. 
    This stage can include substages of prepacking, packing, global placement, legalization, and detailed placement. 
    The objective is to place the logical cell instances of the netlist onto physical BELs on the device in a way that minimizes delay, satisfies design constraints, and simplifies routing.

@@

    \begin{figure*}
        \centering
        \includegraphics[width=\textwidth]{figures/hierarchy.png}
        \caption{Architecture Hierarchy of a Xilinx FPGA (Ultrascale)}
        \label{fig:hierarchy}
    \end{figure*}

@@

\section{RapidWright API}
    \begin{figure}
        \centering
        \includegraphics[width=6.5cm]{figures/vivado_dcps.png}
        \caption{Showing how RapidWright can take a design checkpoint at different stages of the Vivado toolchain, make customizations, then feed it back into the toolchain.}
        \label{fig:vivado_dcps}
    \end{figure}

    We use the RapidWright API to implement a custom placer for Xilinx 7-Series FPGAs.
    RapidWright is an open source Java framework that enables netlist implementation manipulation for modern Xilinx FPGAs. 
    Figure \ref{fig:vivado_dcps} shows how the API can take design checkpoints in the form of .dcp files generated at any stage of the design implementation, allow the user to make custom optimizations, and reintegrate it back into the design flow. 
    This allows the user to substitute any stage in the default FPGA design flow offered by Vivado with user-defined optimizations and implementations to meet complex design requirements. 
    There are three packages of objects that are most relavent to building our placer: the edif package, design package, and device package. 
    RapidWright reads in a post-synthesis design checkpoint and allows us to traverse the netlist's logical EDIFCells through various function calls. 
    For each EDIFCell object in the netlist, we create a Cell design object, which we then place onto physical device BEL object. 
    CellPins and Nets are automatically generated from EDIFHierPortInst and EDIFHierNet upon the Cell's creation from an EDIFHierCellInst.

    These three steps correspond to our placement substages of PrePacking, Packing, and Placement, respectively.
    Figure \ref{fig:edif_design_device} illustrates the flow between these three substages in relation to the RapidWright object classes.

    \begin{figure*}
        \centering
        \includegraphics[width=\textwidth]{figures/edif_design_device.png}
        \caption{The placement flow illustrating PrePacking into Packing into Placement for a logical netlist with 3 EDIF cells. EHCI is short for EDIFHierCellInst. EHPI is short for EDIFHierPortInst.}
        \label{edif_design_device}
    \end{figure*}

    For a complete understanding of the RapidWright packages, please refer to the short API Overview \url{https://www.rapidwright.io/docs/RapidWright_Overview.html}.
    Listed below are the most relevant objects to our placer and a short description.

    \begin{figure*}
        \centering
        \includegraphics[width=\textwidth]{figures/edif_netlist.png}
        \caption{Examples of edif netlist objects for a Xilinx FPGA}
        \label{fig:routing_objects}
    \end{figure*}

    \begin{figure*}
        \centering
        \includegraphics[width=\textwidth]{figures/routing_objects.png}
        \caption{Examples of device routing objects in a Xilinx FPGA}
        \label{fig:routing_objects}
    \end{figure*}

    \subsection{EDIF Objects}
        \begin{itemize}
            \item EDIFNetlist
            \item EDIFHierNet
            \item EDIFHierCellInst
            \item EDIFHierPortInst
        \end{itemize}
    \subsection{Design Objects}
        \begin{itemize}
            \item Netlist
            \item Net
            \item Cell
            \item CellPin
            \item SiteInst
            \item SitePinInst
        \end{itemize}
    \subsection{Device Objects}
        \begin{itemize}
            \item BEL
            \item PIP (programmable interconnect point)
            \item Site
            \item SitePin
            \item Tile
            \item ClockRegion
        \end{itemize}

    In Vivado, all designs post synthesis have a logical netlist that can be exported as an Electronic Design Interchange Format (EDIF) file or within a .dcp checkpoint file. 
    This is the result of the synthesis tool parsing the raw Verilog source files, inferring logical behavior, and instantiating primitive cells to emulate the desired logical behavior. 
    The final stage is often referred to as "technology mapping", as the primitive cells are specific to the FPGA architecture and consequently the FPGA manufacturer.

    In its current state, RapidWright has several SA-based "BlockPlacers", a general Router, but no general Placer. 
    The BlockPlacer takes multiple pre-implemented designs (that is, designs that have been placed and routed), and instantiates them as hard macros in the form of ModuleInsts. 
    ModuleInsts are a collection of SiteInsts whose positional relationship to eachother are fixed. 
    The BlockPlacer takes these hard ModuleInsts and performs simulated annealing on them to find an optimal placement. 
    The Router takes the placed design and routes the nets onto the device wires and programmable interconnects.

    To fill this gap, we attempt to build a general Placer that can take a raw EDIF netlist and create a placed design. 
    As mentioned before, our placer will be divided into three substages: prepacking, packing, and placement.
    We use this API to implement a crude general placer, in our case, a simulated annealer. 
    (Show what parts of the FPGA toolchain that RapidWright already has and how our placer fits into that toolchain).

\section{Placement Implementation}
    (Describe the algorithm, then in context of 7-Series arhictecture). \\

    \subsection{PrePacking}
        In the prepacking stage, we traverse the raw EDIF Netlist to identify recurring cell patterns. 
        In particular, we will specifically look for CARRY chains, DSP cascades, and LUT-FF pairs as these patterns are most prevalent in all FPGA designs.
        \subsection{CARRY Chains}
            Each SLICE Site contains one CARRY4 BEL.
            Each CLB Tile contains two SLICE Sites.
            CARRY chains span vertically across multiple SLICE Sites/Tiles. (Show picture)
            \begin{figure}
                \centering
                \includegraphics[width=3.0cm]{figures/carry_chain_routes.png}
                \includegraphics[width=3.0cm]{figures/dsp_cascade_routes.png}
                \caption{A Carry Chain of length 6 spanning 6 Tiles. On the left, the cell-only representation. On the right, cells with routing representation.}
                \label{fig:device_carry_chain_routing}
            \end{figure}
        \subsection{DSP Cascades}
            Each DSP Site contains one DSP BEL.
            Each DSP Tile contains two DSP Sites.
            DSP chains can span vertically across multiple DSP Sites/Tiles.

        \subsection{LUT-FF Pairs}
            Identify unique CE-SR net pairs. All FF cells in a Site must share the same CE and SR nets.
    \subsection{Packing}
        In the packing stage, we take the identified cell clusters and pack them into SiteInst Design objects which target the Device Site objects.
        \subsection{CLB Sites}
            Can support LUT-FF pairs, loose LUTs, loose FFs, CARRY chains.
            Each SLICE has 8 "lanes" of LUT-FFs. 4 LUT5s and 4 LUT6s. 8 FFs.
            For SLICEMs, LUT6s can be configured as shallow 32-bit LUTRAMs or "RAMS32".
    \subsection{Placement}
        Create BEL "fields": CLB, DSP, CLB and DSP Chains, RAMs

@@

>>>>

\documentclass[twocolumn]{article}
\usepackage{hyperref}
\usepackage{enumitem}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{mathpazo}
\usepackage{listings}
% \usepackage{multicol}
\usepackage[a4paper,
            bindingoffset=0.2in,
            left=1in,
            right=1in,
            top=1in,
            bottom=1in,
            footskip=.25in]{geometry}

\lstset{
  basicstyle=\ttfamily\small,
  breaklines=true,
  columns=fullflexible,
  frame=single,
  backgroundcolor=\color{gray!10},
  captionpos=b
}

@@

    This brings us to modern day FPGA architectures.
    To meet the needs of increasingly complex designs, FPGA vendors introduced heterogeneous FPGAs. 
    In these devices, hard macros such as Block RAM (BRAM) and Digital Signal Processing (DSP) slices are integrated into the programmable logic fabric along with CLBs. 
    This design enables the direct instantiation of common subsystems like memories and multipliers, without having to recreate them from scratch using CLBs. 
    Major vendors such as Xilinx and Altera now employ heterogeneous island-style architectures in their devices. 
    As designs become increasingly large and complex, FPGAs become increasingly heterogenous by incorporating a wider variety of hard macros into the fabric.

@@

    \begin{figure}
        \centering
        \includegraphics[width=7.0cm]{figures/design_flow.png}
        \caption{A typical FPGA design and verification workflow}
        \label{fig:design_flow}
    \end{figure}

    Figure \ref{fig:design_flow} shows a typical FPGA design and verification workflow.
    The FPGA design flow takes a high-level digital design written in an HDL and generates a configuration file that programs the physical FPGA.

@@

    Next, the \textbf{synthesis} tool interprets the HDL code and generates a netlist that represents the design in terms of primitive cells, such as lookup tables (LUTs) and flip flops. 
    This is the result of the synthesis tool parsing the raw Verilog source files, inferring logical behavior, and instantiating primitive cells to emulate the desired logical behavior. 
    The final stage is often referred to as "technology mapping", as the primitive cells are specific to the FPGA architecture and consequently the FPGA manufacturer.

    After synthesis, the \textbf{placement} tool assigns each logic element in the netlist a specific location within the FPGA’s fabric. 
    Depending on the placement strategy, this stage can include substages of prepacking, packing, global placement, legalization, and detailed placement. 
    Our placer will only involve prepacking, packing, and detailed placement for simplicity. 
    The objective is to place the logical cell instances of the netlist onto physical BELs on the device in a way that minimizes delay, satisfies design constraints, and simplifies routing.

@@

    \begin{figure*}[!hbt]
        \centering
        \includegraphics[width=\textwidth]{figures/hierarchy.png}
        \caption{Architecture Hierarchy of a Xilinx FPGA (Ultrascale)}
        \label{fig:hierarchy}
    \end{figure*}

@@

\section{RapidWright API}
    \begin{figure}
        \centering
        \includegraphics[width=6.0cm]{figures/vivado_dcps.png}
        \caption{Design flow showing how RapidWright can take a design checkpoint, allow user customization, and reintegrate it back into the design flow.}
        \label{fig:vivado_dcps}
    \end{figure}

    \begin{figure*}[!htt]
        \centering
        \includegraphics[width=\textwidth]{figures/edif_design_device.png}
        \caption{The placement flow illustrating PrePacking into Packing into Placement for a logical netlist with 3 EDIF cells. EHCI is short for EDIFHierCellInst. EHPI is short for EDIFHierPortInst.}
        \label{fig:edif_design_device}
    \end{figure*}

    We use the RapidWright API to implement a custom placer for Xilinx 7-Series FPGAs.
    RapidWright is an open source Java framework that enables netlist implementation manipulation for modern Xilinx FPGAs. 
    Figure \ref{fig:vivado_dcps} shows how the API can take design checkpoints in the form of .dcp files generated at any stage of the design implementation, allow the user to make custom optimizations, and reintegrate it back into the design flow. 
    This allows the user to substitute any stage in the default FPGA design flow offered by Vivado with user-defined optimizations and implementations to meet complex design requirements. 
    There are three packages of objects that are most relavent to building our placer: the \textbf{edif} package, \textbf{design} package, and \textbf{device} package. 
    Figure \ref{fig:routing_objects} shows the device routing objects as shown in the Vivado device viewer while figure \ref{fig:edif_netlist} shows the EDIF netlist objects as shown in the Vivado netlist viewer.
    Figure \ref{fig:edif_design_device} illustrates the flow between these packages in relation to our PrePacking, Packing, and Placement flow.

    RapidWright reads in a post-synthesis design checkpoint and allows us to traverse the netlist's logical EDIFCells through various function calls. 
    For each EDIFCell object in the netlist, we create a Cell design object, which we then place onto physical device BEL object. 
    CellPins and Nets are automatically generated from EDIFHierPortInst and EDIFHierNet upon the Cell's creation from an EDIFHierCellInst.

    These three steps correspond to our placement substages of PrePacking, Packing, and Placement, respectively.

    For a complete understanding of the RapidWright packages, please refer to the short API Overview \url{https://www.rapidwright.io/docs/RapidWright_Overview.html}.

    \begin{figure*}[!htt]
        \centering
        \includegraphics[width=12.0cm]{figures/edif_netlist.png}
        \caption{Examples of edif netlist objects for a Xilinx FPGA}
        \label{fig:edif_netlist}
    \end{figure*}

    \begin{figure*}[!htt]
        \centering
        \includegraphics[width=\textwidth]{figures/routing_objects.png}
        \caption{Examples of device routing objects in a Xilinx FPGA}
        \label{fig:routing_objects}
    \end{figure*}

\section{Placement Implementation}

    With a basic understanding of FPGA architecture, design placement, and RapidWright, we can begin implementing our SA placer. 
    Here we outline in detail each substage of our implementation.

    \subsection{PrePacking}
        In the prepacking stage, we traverse the raw EDIF Netlist to identify recurring Cell patterns. 
        We do this because there are certain desirable cell configurations that should be clustered together to exploit some proximity characteristics of the FPGA architecture 
        There are also some cell configurations that must necessarily be placed in certain ways in relation to each other to ensure legality within the architecture constraints. 
        First, we traverse the entire EDIF netlist and identify all unique Cell types and group them together via a {\tt HashMap<String, List<EDIFHierCellInst>>}, where String is the name cell type. 
        Cell type groups can include {\tt CARRY4}, {\tt LUT1-6}, {\tt FF}, {\tt DSP48E1}, {\tt RAMB18E1}, and others. 
        We specifically look for CARRY chains, DSP cascades, and LUT-FF pairs as these patterns are most prevalent in all FPGA designs.

        \subsection{CARRY Chains}
            To find CARRY chains, we take the CARRY4 group and follow the below pseudocode. 
            \begin{verbatim}
                for (EDIFHierCellInst ehci : CARRYCells) {
                    ehci.getPorts()

                }
            \end{verbatim}
            Each SLICE Site contains one CARRY4 BEL.
            Each CLB Tile contains two SLICE Sites.
            CARRY chains span vertically across multiple SLICE Sites/Tiles. (Show picture)
            \begin{figure}
                \centering
                \includegraphics[width=3.0cm]{figures/carry_chain_routes.png}
                \includegraphics[width=3.0cm]{figures/dsp_cascade_routes.png}
                \caption{A Carry Chain of length 6 spanning 6 Tiles. On the left, the cell-only representation. On the right, cells with routing representation.}
                \label{fig:device_carry_chain_routing}
            \end{figure}
        \subsection{DSP Cascades}
            Each DSP Site contains one DSP BEL.
            Each DSP Tile contains two DSP Sites.
            DSP chains can span vertically across multiple DSP Sites/Tiles.

        \subsection{LUT-FF Pairs}
            Identify unique CE-SR net pairs. All FF cells in a Site must share the same CE and SR nets.
    \subsection{Packing}
        In the packing stage, we take the identified cell clusters and pack them into SiteInst Design objects which target the Device Site objects.
        \subsection{CLB Sites}
            Can support LUT-FF pairs, loose LUTs, loose FFs, CARRY chains.
            Each SLICE has 8 "lanes" of LUT-FFs. 4 LUT5s and 4 LUT6s. 8 FFs.
            For SLICEMs, LUT6s can be configured as shallow 32-bit LUTRAMs or "RAMS32".
    \subsection{Placement: Simulated Annealing}
        Create BEL "fields": CLB, DSP, CLB and DSP Chains, RAMs

@@

\section{Placement Results}
